A White Noise Approach to Evolutionary Ecology

### 2020-04-18 .A White Noise Approach To Evolutionary Ecology/draft_04_18_20.Rmd
final draft

---
title: "A White Noise Approach to Evolutionary Ecology"
author: "Bob Week, Steve Krone, Scott L. Nuismer, Luke J. Harmon"
date: "3/1/2020"
abstract: "We derive the dynamics of the distribution of a quantitative character and the abundance of a biological population from a stochastic partial differential equation driven by space-time white noise. In the process we develop a useful set of heuristics to operationalize the powerful, but abstract theory of white noise and measure-valued stochastic processes. This approach allows us to compute the full implications of demographic stochasticity on phenotypic distributions and abundances of populations. We demonstrate the utility of our approach by deriving a quantitative genetic model of diffuse coevolution mediated by exploitative competition for a continuum of resources. Other than trait and abundance distributions, this model predicts interaction networks parameterized by rates of interactions, competition coefficients, and selection gradients. We briefly investigate the relationship between selection gradients and competition coefficients. This illustrative investigation suggests selection gradients can be either positively or negatively correlated with competition coefficients depending on the ratio of interspecific trait variation to intraspecific trait variation. Hence, this approach contributes to the development of a synthetic theory of evolutionary ecology by formalizing first principle derivations of stochastic models that underlie rigorous investigations of the relationship between feedbacks of biological processes and the patterns of diversity they produce."
bibliography: ref.bib
output:
  pdf_document: 
    extra_dependencies:
      - float
      - flafter
    fig_caption: yes
    fig_height: 3
    fig_width: 4
    keep_tex: yes
    number_sections: yes
header-includes:
  - \usepackage{mathtools}
  - \usepackage{mathrsfs}
  - \usepackage{hyperref}
  - \usepackage[left]{lineno}
  - \usepackage{amsmath}
  - \linenumbers
editor_options: 
  chunk_output_type: console
---

@@

Current theoretical approaches to synthesize evolution and ecology have capitalized on the fact that biological fitness plays a key role in determining both sets of dynamics. While correlation of fitness and genotype is the basis of evolution by natural selection, the mean fitness across all individuals in a population determines the growth, stasis or decline of abundance. In section \ref{deterministic} we review the mathematical formalization of this connection, which has been established in the contexts of population genetics [@9781932846126; @joanroughgarden1979], evolutionary game theory [@Hofbauer1998; @9780674023383; @Lion2018], quantitative genetics [@Lande1982; @Doebeli1996; @Lion2018] and a unifying framework for these three distinct approaches to evolutionary theory [@Champagnat2006] which includes our approach as a special case. We note here there are two primary differences between our aims and those of Champagnat, Ferrière, and Méléard [-@Champagnat2006]. First, instead of developing a unifying approach to evolutionary theory, we focus on developing a stochastic synthesis of population dynamics and quantitative genetics including specific expressions for the dynamics of abundance, mean trait and additive genetic variance. Second, although Champagnat, Ferrière, and Méléard [-@Champagnat2006] provide rigorous proofs of results relevant to our approach, they do not communicate the nessecary tools we make use of for deriving the dynamics of abundance, mean trait and additive genetic variance. We aim to communicate these tools to a wide audience by providing an intuitive, yet accurate treatment of white noise and associated heuristics useful for deriving new stochastic models of evolutionary ecology.

Reviewing the above accomplishments reveals a beautiful synthesis of evolution and population ecology. However, it also reveals a gap in theoretical approaches to incorporate the intrinsically random nature of populations. Specifically, in theoretical quantitative genetics the derivation of a population's response to random genetic drift is derived in discrete time under the assumption of constant effective population size using arguments based on properties of random samples [@Lande1976]. Though this approach conveniently mimics the formalism provided by the Wright-Fisher model of population genetics, real population sizes fluctuate over time. Furthermore, since these fluctuations are themselves stochastic, it seems natural to derive expressions for the evolutionary response to demographic stochasticity and consider how the results relate to characterizations of random genetic drift. This can be done in continuous time for population genetic models without too much technical overhead, assuming a finite number of alleles [@Lande2009; @Parsons2010; @Gomulkiewicz2017]. However, for populations with a continuum of types, such as a quantitative trait, finding a formal approach to derive the evolutionary response to demographic stochasticity has remained a vexing mathematical challenge. Previous work integrating stochastic processes with quantitative genetic models have profited from the simplifying assumption that phenotypic and genetic variances remain constant over time, yeilding analytical approximations for the joint dynamics of abundance and the mean of a quantitative trait [@Engen2013]. These results not only ignore the evolution genetic variance, but they focus on environmental stochasticity and assume sufficiently large populations so that demographic stochasticity is negligible. Here we account for both demographic stochasticity and the evolution of additive genetic variance by combining the calculus of white noise with results on rescaled limits of branching Brownian motion processes (BBM) and stochastic partial differential equations (SPDE), which are stochastic analogs of deterministic partial differential equations (PDE). Our goal has two components: 1) Establish a novel synthetic approach to theoretical evolutionary ecology that provides a formal connection between demographic stochasticity and random genetic drift in the context of quantitative traits. 2) Communicate some useful properties of space-time white noise, BBM and SPDE to as wide of audience as possible. With this goal in mind we will not provide a rigorous treatment of any of these deep subjects. Instead, we introduce a set of heuristics that only require the basic concepts of Reimann integration, partial differentiation and some exposure to Brownian motion and stochastic ordinary differential equations (SDE). For a concise introduction to SDE and Brownian motion, we recommend the primer by Evans [-@9781470410544]. Rigorous treatments of SPDE and rescaled limits of BBM can be found in Walsh [-@Walsh] and Dawson [-@dawson1993measure] respectively.

Through accomplishing these goals this work yields five main results. Firstly, we develop a concise, but accurate approach for understanding SPDE and provide heuristics to empower mathematical evolutionary ecologists to derive novel models that relax key assumptions commonly made in combining models of evolutionary and abundance dynamics. Second, we formally derive general SDE that express the joint dynamics of abundance, mean trait and additive genetic variance for populations exhibiting a wide range of potential phenotypic distributions. Third, we simplify these SDE by approximating the abundance density of phenotypes with a Gaussian curve and integrate the results with a model of imperfect inheritance to find formula that generalize classical quantitative genetic expressions. This includes an expression for additive genetic variance analogous to the expression for the response in mean phenotype found by Lande [-@Lande1976]. Fourth, following classical niche theory we provide a rigorous mechanistic derivation of fitness for populations experiencing resource competition and abiotic stabilizing selection using diffusion limits of measure-valued branching processes. This calculation provides a demonstration for performing such diffusion limits that can be applied in other contexts. Fifth, we explore implications of the resulting model of diffuse coevolution by investigating the relationship between selection gradients and competition coefficients.

We now summarize the basic structure of the paper. To provide motivation for the stochastic equations developed later and background for our model of coevolution, we begin with \S\ref{deterministic} by briefly summarizing derivations of deterministic dynamics of biological populations. Starting with a PDE (which we refer to as the Deterministic Asexual Gaussian allelic model with Abundance dynamics, abbreviated DAGA), we arrive at a general set of ordinary differential equations modelling the dynamics of abundance, trait mean and trait variance. We discuss the relationship between DAGA and previous models of phenotypic evolution that have been employed in theoretical quantitative genetics. Noting that a natural stochastic generalization of DAGA has been previously derived, we find a path to derive SDE describing the evolutionary response to demographic stochasticity. We accomplish this in \S\ref{wnc} by discussing how a diffusion limit of a spatially structured branching process (i.e., a measure-valued branching process) leads to the natural SPDE appropriate for our study. We refer to this SPDE as the Stochastic Asexual Gaussian allelic model with Abundance dynamics (abbreviated SAGA). The diffusion limit in turn provides a rigorous method for constructing fitness functions used in models of evolutionary ecology. However, before we discuss measure-valued braching processes and their diffusion limits, we find it necessary to review some fundamental results on the calculus of white noise and introducing a set of mathematical tools based on these results. In \S\ref{wnc} and SM \S\ref{SDE_DERIV} we employ these tools to derive a system of SDE generalizing our deterministic results to account for demographic stochasticity. We find that the effect of demographic stochasticity on the evolution of mean trait and phenotypic variance characterizes the process of random genetic drift in continuous time. However, although biologically insightful, these equations remain difficult to analyze and implement numerically. In \S\ref{equations} we use an assumption of normally distributed trait values to simplify these expressions into formulae that are much easier to work with. We then account for the constraint of adaptive evolution on the amount of heritable variation in a population by extending these results via a model of imperfect inheritance. The resulting equations coincide with classical results in quantitative genetics as a special case. In \S\ref{coev} we combine the derived equations of population dynamics with classical niche theory to formulate a model of coevolution across a guild of $S$ species participating in exploitative competition along a common resource continuum. In SM \S\ref{diffuse} we apply a classical theorem on rescaled limits of BBM that allow for ecological interactions to provide a rigorous derivation. To gain biological insight, in \S\ref{dynamics} we numerically integrate our model of coevolution for $S=100$ species, tracking the dynamics of traits and abundances under scenarios of weak and strong competition. We include an account of the natural history of the simulated system and discuss the significance of demographic stochasticity for structuring ecological communities. In \S\ref{ecoevo} we provide expressions for selection gradients and competition coefficients implied by our model and use these expressions to investigate the relationship between the degree of competition and strength of coevolution. Finally, \S\ref{conclusion} concludes with a summary of accomplishments, a few remarks on the limits of this approach and future directions to incorporate more explicitly the effects of small populations, the genetic architecture of quantitative traits, feedbacks with ecosystem processes and the macroevolutionary history of interacting lineages.

@@

At the core of our approach is a stochastic analog of the replicator equation with mutation in continuous time and phenotypic space [@Taylor1978; @Schuster1983]. From this stochastic replicator-mutator equation we derive a system of SDE for the dynamics of abundance, mean trait and additive genetic variance of a population. Hence, our approach develops a quantitative genetic theory of evolutionary ecology. A popular alternative to quantitative genetics is the theory of adaptive dynamics. As demonstrated by Page and Nowak [-@PAGE2002] and Champagnat, Ferrière, and Méléard [-@Champagnat2006], the canonical equation of adaptive dynamics can be derived from the replicator-mutator equation. Thus, one could start from the atomic roots of our approach and pursue a stochastic adaptive dynamic theory instead. We choose the former in anticipation of an extension of our approach that explicitly models the genetic details of populations.

In this section we review the derivations of the replicator-mutator equation and trait dynamics from abundance dynamics and extend these formulae along with related results to the stochastic case. The results established in this section provide the framework from which larger scale ecological stuctures, such as species abundance distributions and interaction networks, can be computed.

@@

Our review begins by considering the dynamics of an asexually reproducing population in a homogeneous environment. For simplicity, we first assume individuals are haploid and carry one of $K$ alleles each with a different fitness expressed as growth rate before introducing a model involving a quantitative trait. Under these assumptions, the derivation of the evolution of allele frequencies due to natural selection can be derived from expressions of exponential growth. This, and a few related approaches, have been provided by Crow and Kimura [-@9781932846126, \S 5.3]. Specifically, denoting $\nu_i$ the abundance of individuals with allele $i$ and $m_i$ the growth rate of allele $i$ [called the Malthusian parameter in @9781932846126], we have
\begin{equation}\label{first}
\frac{d\nu_i}{dt}=m_i\nu_i.
\end{equation}
Starting from this model, we get the total abundance of the population as $N=\sum_{i=1}^K\nu_i$, the frequency of allele $i$ as $p_i=\nu_i/N$ and the mean fitness of the population as $\bar m=\sum_{i=1}^Kp_im_i$. Hence, we can employ some elementary calculus to derive the dynamics of abundance $dN/dt$ and the dynamics of allele frequencies $dp_1/dt,\dots,dp_K/dt$ as
\begin{equation}
\frac{dN}{dt}=\sum_{i=1}^K\nu_im_i=N\sum_{i=1}^Kp_im_i=\bar mN,
\end{equation}
\begin{equation}\label{popgen}
\frac{dp_i}{dt}=\frac{d}{dt}\frac{\nu_i}{N}=\frac{1}{N^2}\left(N\frac{d\nu_i}{dt}-\frac{dN}{dt}\nu_i\right)
=\frac{1}{N}\left(m_i\nu_i-\bar m Np_i\right)=(m_i-\bar m)p_i.
\end{equation}
Two important observations of these equations include 1) mean fitness $\bar m$ determines the abundance dynamics of the entire population and 2) allele $i$ will increase (decrease) in frequency if $m_i>\bar m$ ($<\bar m$). Equation (\ref{popgen}) is known in the field of evolutionary game theory as the replicator equation [@Hofbauer1998; @9780674023383; @Lion2018; @Taylor1978; @Schuster1983]. Instead of being explicitly focused on alleles, the replicator equation describes the fluctuations of relative abundances of various _types_ in a population in terms of the vital rates of each type. Using a matrix of transition rates between differing types, it is straight-forward to extend the replicator equation to include mutation, which is known as the replicator-mutator equation [@9780674023383].

@@

To stay within the realm of biological plausibility we require a set technical assumptions. First, we assume the initial abundance density is continuous, non-negative, integrable and has finite trait mean and variance. That is, we assume $\nu(x,0)$ is continuous in $x$, satisfies $\nu(x,0)\geq0$ for all $x\in\mathbb{R}$ and
\begin{equation}
N(0)=\int_{-\infty}^{+\infty}\nu(x,0)dx<+\infty,
\end{equation}
\begin{equation}
-\infty<\bar x(0)=\int_{-\infty}^{+\infty}xp(x,0)dx<+\infty,
\end{equation}
\begin{equation}
\sigma^2(0)=\int_{-\infty}^{+\infty}(x-\bar x(0))^2p(x,0)dx<+\infty,
\end{equation}
where $\bar x(t)$ and $\sigma^2(t)$ are respectively the mean trait and phenotypic variance at time $t\geq0$. Second, we assume selection is determined by the growth rate $m(h,x)$ that is differentiable with respect to both arguments and is bounded above by some $r\in\mathbb{R}$ for all $x\in\mathbb{R}$ and $h\geq0$. We denote by $K$ an operator that takes functions such as $\nu(x,t)$ as its argument and returns other functions that may still depend on the spatial variable $x$ and the temporal variable $t$. For our application we will be focused on operators $K$ that can be written as $(K\nu)(x,t)=\int_{-\infty}^{+\infty}\kappa(x-y)\nu(y,t)dy$ for some non-negative and bounded function $\kappa$. We set $h=(K\nu)(x,t)$ to account for nonlocal effects of the abundance density $\nu(x,t)$ on the fitness of individuals with trait value $x$ [@Volpert2014]. As an example of a nonlocal effect, consider a population experiencing intraspecific competition for a common resource such that the competitive interactions are not mediated by any set of traits. In this case all individuals in the population effect the fitness of one another even though they may exhibit a diverse range of trait values. The term nonlocal comes from interpreting the trait value as spatial location. From here on we abbreviate $m(h,x)=m\big((K\nu)(x,t),x\big)$ to just $m(\nu,x)$. Third, we assume mutation is captured by diffusion with coefficient $\frac{\mu}{2}$. With these technicalities addressed we model the demographic dynamics of a population and the dynamics of a quantitative character simultaneously by the PDE
\begin{equation}\label{eq1}
\frac{\partial}{\partial t}\nu(x,t)=m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)
\end{equation}
with the initial condition $\nu(x,0)$ described above. This PDE is semilinear due to the dependency of the growth rate $m(\nu,x)$ on the solution $\nu(x,t)$ and is referred to as a scalar reaction-diffusion equation [@zheng2004nonlinear; @lawrenceevans2010]. To ensure existence and uniqueness of solutions to (\ref{eq1}), we further assume that for each $M>0$ there exists a constant $L_M>0$ depending on $M$ such that $\int_\mathbb{R}\nu_1(x)dx,\int_\mathbb{R}\nu_2(x)dx\leq M$ implies
\begin{equation}\label{local_lipschitz}
\int_\mathbb{R}\left|m(\nu_1,x)\nu_1(x)-m(\nu_2,x)\nu_2(x)\right|dx\leq L_M\int_\mathbb{R}\nu_1(x)-\nu_2(x)dx.
\end{equation}
In SM \S\ref{finite} we combine the above assumptions on growth rate, mutation and initial condition to prove that solutions to (\ref{eq1}) satisfy $N(t),|\bar x(t)|,\sigma^2(t)<+\infty$ for all $t\geq0$.

>>>>

---
title: "A White Noise Approach to Evolutionary Ecology"
author: "Bob Week, Steve Krone, Scott L. Nuismer, Luke J. Harmon"
date: "3/1/2020"
abstract: "We derive the dynamics of the distribution of a quantitative character and the abundance of a biological population from a stochastic partial differential equation driven by space-time white noise. In the process we develop a useful set of heuristics to operationalize the powerful, but abstract theory of white noise and measure-valued stochastic processes. This approach allows us to compute the full implications of demographic stochasticity on phenotypic distributions and abundances of populations. We demonstrate the utility of our approach by deriving a quantitative genetic model of diffuse coevolution mediated by exploitative competition for a continuum of resources. Other than trait and abundance distributions, this model predicts interaction networks parameterized by rates of interactions, competition coefficients, and selection gradients. We briefly investigate the relationship between selection gradients and competition coefficients. This illustrative investigation suggests absolute values of linear selection gradients are negatively correlated with competition coefficients when the ratio of interspecific trait variation to intraspecific trait variation is approximately less than $1.25$ and positively correlated otherwise. The inverse relationship holds for quadratic selection gradients except with threshold diversity ratio equal to unity instead of $1.25$. Hence, this approach contributes to the development of a synthetic theory of evolutionary ecology by formalizing first principle derivations of stochastic models that underlie rigorous investigations of the relationship between feedbacks of biological processes and the patterns of diversity they produce."
bibliography: ref.bib
output:
  pdf_document: 
    extra_dependencies:
      - float
      - flafter
    fig_caption: yes
    fig_height: 3
    fig_width: 4
    keep_tex: yes
    number_sections: yes
header-includes:
  - \usepackage{mathtools}
  - \usepackage{mathrsfs}
  - \usepackage{hyperref}
  - \usepackage[left]{lineno}
  - \usepackage{amsmath}
  - \linenumbers
editor_options: 
  chunk_output_type: console
---

@@

Current theoretical approaches to synthesize evolution and ecology have capitalized on the fact that biological fitness plays a key role in determining both sets of dynamics. While correlation of fitness and genotype is the basis of evolution by natural selection, the mean fitness across all individuals in a population determines the growth, stasis or decline of abundance. In section \ref{deterministic} we review the mathematical formalization of this connection, which has been established in the contexts of population genetics [@9781932846126; @joanroughgarden1979], evolutionary game theory [@Hofbauer1998; @9780674023383; @Lion2018], quantitative genetics [@Lande1982; @Doebeli1996; @Lion2018] and a unifying framework for these three distinct approaches to evolutionary theory [@Champagnat2006] which includes our approach as a special case. We note here there are two primary differences between the results presented in this paper and those of Champagnat, Ferrière and Méléard [-@Champagnat2006]. First, instead of developing a unifying approach to evolutionary theory, we focus on developing a stochastic synthesis of population dynamics and quantitative genetics including specific expressions for the dynamics of abundance, mean trait and additive genetic variance. Second, although Champagnat, Ferrière and Méléard [-@Champagnat2006] provide rigorous proofs of results relevant to our approach, they do not communicate the nessecary tools we make use of for deriving the dynamics of abundance, mean trait and additive genetic variance. We aim to communicate these tools to a wide audience by striking a balance between intuitive appeal and mathematical rigor.

Although reviewing previously established theoretical results reveals a beautiful synthesis of evolution and population ecology, it also reveals a gap in theoretical approaches to incorporate the intrinsically random nature of populations. Specifically, in theoretical quantitative genetics the derivation of a population's response to random genetic drift is derived in discrete time under the assumption of constant effective population size using arguments based on properties of random samples [@Lande1976]. Though this approach conveniently mimics the formalism provided by the Wright-Fisher model of population genetics, real population sizes fluctuate over time. Furthermore, since these fluctuations are themselves stochastic, it seems natural to derive expressions for the evolutionary response to demographic stochasticity and consider how the results relate to characterizations of random genetic drift. This can be done in continuous time for population genetic models without too much technical overhead, assuming a finite number of alleles [@Lande2009; @Parsons2010; @Gomulkiewicz2017]. However, for populations with a continuum of types, such as a quantitative trait, finding a formal approach to derive the evolutionary response to demographic stochasticity has remained a vexing mathematical challenge. In this paper we close this gap by combining the calculus of white noise with results on rescaled limits of branching Brownian motion processes (BBM) and stochastic partial differential equations (SPDE), which are stochastic analogs of deterministic partial differential equations (PDE). Our goal has two components: 1) Establish a novel synthetic approach to theoretical evolutionary ecology that provides a formal connection between demographic stochasticity and random genetic drift in the context of quantitative traits. 2) Communicate some useful properties of space-time white noise, BBM and SPDE to as wide of audience as possible. With this goal in mind we will not provide a rigorous treatment of any of these deep subjects. Instead, we introduce a set of heuristics that only require the basic concepts of Reimann integration, partial differentiation and some exposure to Brownian motion and stochastic ordinary differential equations (SDE). For a concise introduction to SDE and Brownian motion, we recommend the primer by Evans [-@9781470410544]. Rigorous treatments of SPDE and rescaled limits of BBM can be found in Walsh [-@Walsh] and Dawson [-@dawson1993measure] respectively.

Through accomplishing these goals this work yields five main results. Firstly, we develop a concise, but accurate approach for understanding SPDE and provide heuristics to empower mathematical evolutionary ecologists to derive novel models that relax key assumptions commonly made in combining models of evolutionary and abundance dynamics. Second, we formally derive general SDE driven by demographic stochasticity that express the joint dynamics of abundance, mean trait and additive genetic variance for populations exhibiting a wide range of potential phenotypic distributions. Although these dynamics have been considered before while accounting for demographic stochasticity [@Polechov2015], to our knowledge this is the first time they have been rigorously derived from first principles. Third, we simplify these SDE by approximating the abundance density of phenotypes with a Gaussian curve and integrate the results with a model of imperfect inheritance to find formula that generalize classical quantitative genetic expressions. This includes an expression for the response of additive genetic variance to selection in terms of fitness gradients analogous to the expression for the response in mean phenotype found by Lande [-@Lande1976]. Fourth, following classical niche theory we provide a rigorous mechanistic derivation of fitness for populations experiencing resource competition and abiotic stabilizing selection using diffusion limits of measure-valued branching processes. This calculation provides a demonstration for performing such diffusion limits that can be applied in other contexts. Fifth, we explore implications of the resulting model of diffuse coevolution by investigating the relationship between selection gradients and competition coefficients, discovering the conditions under which these quantities positively or negatively covary.

We now summarize the basic structure of the paper. To provide motivation for the stochastic equations developed later and background for our model of coevolution, we begin with \S\ref{deterministic} by briefly summarizing derivations of deterministic evolutionary and demographic dynamics. Starting with a PDE (which we refer to as the Deterministic Asexual Gaussian allelic model with Abundance dynamics, abbreviated DAGA), we arrive at a general set of ordinary differential equations modelling the dynamics of abundance, trait mean and trait variance. We discuss the relationship between DAGA and previous models of phenotypic evolution that have been employed in theoretical quantitative genetics. In \S\ref{wnc} we discuss how a diffusion limit of a spatially structured branching process (i.e., a measure-valued branching process) leads to a SPDE that provides a natural stochastic generalization of DAGA. We refer to this SPDE as the Stochastic Asexual Gaussian allelic model with Abundance dynamics (abbreviated SAGA). The diffusion limit in turn provides a rigorous method for constructing fitness functions used in models of evolutionary ecology. However, before we discuss measure-valued braching processes and their diffusion limits, we find it necessary to review some fundamental results on the calculus of white noise and introduce a set of mathematical tools based on these results. In SM \S\ref{SDE_DERIV} we employ these tools to derive a system of SDE generalizing our deterministic results to account for demographic stochasticity. We find the effect of demographic stochasticity on the evolution of mean trait and phenotypic variance characterizes the process of random genetic drift in continuous time. However, although biologically insightful, these equations remain difficult to analyze and implement numerically. In \S\ref{equations} we use an assumption of normally distributed trait values to simplify these expressions into tractable formula that express the evolutionary response to selection in terms of fitness gradients. We then account for the constraint of adaptive evolution on the amount of heritable variation available in a population by extending these results via a model of imperfect inheritance. The resulting equations coincide with classical results in quantitative genetics as a special case. In \S\ref{coev} we combine the derived equations of population dynamics with classical niche theory to formulate a model of coevolution across a guild of $S$ species participating in exploitative competition along a common resource continuum. In SM \S\ref{diffuse} we apply a result on rescaled limits of BBM to provide a rigorous derivation. In \S\ref{dynamics} we illustrate our model of coevolution through numerical integration for $S=100$ species, tracking the dynamics of traits and abundances under scenarios of weak and strong competition. We include an account of the natural history of the simulated system and discuss the significance of demographic stochasticity for structuring ecological communities. In \S\ref{ecoevo} we provide expressions for selection gradients and competition coefficients implied by our model and use these expressions to investigate the relationship between the degree of competition and strength of coevolution. Finally, \S\ref{conclusion} concludes with a summary of accomplishments, a few remarks on the limits of this approach and future directions to incorporate more explicitly the effects of small populations, the genetic architecture of quantitative traits, feedbacks with ecosystem processes and the macroevolutionary history of interacting lineages.

@@

At the core of our approach is a stochastic analog of the replicator equation with mutation in continuous time and phenotypic space [@Taylor1978; @Schuster1983]. From this stochastic replicator-mutator equation we derive a system of SDE for the dynamics of abundance, mean trait and additive genetic variance of a population. Hence, our approach develops a quantitative genetic theory of evolutionary ecology. A popular alternative to quantitative genetics is the theory of adaptive dynamics. As demonstrated by Page and Nowak [-@PAGE2002] and Champagnat, Ferrière and Méléard [-@Champagnat2006], the canonical equation of adaptive dynamics can be derived from the replicator-mutator equation. Thus, one could start from the atomic roots of our approach and pursue a stochastic adaptive dynamic theory instead. We choose the former in anticipation of an extension of our approach that explicitly models the genetic details of populations.

In this section we review the derivations of the replicator-mutator equation and trait dynamics from abundance dynamics and extend these formula along with related results to the stochastic case. The results established in this section provide the framework from which larger scale ecological stuctures, such as species abundance distributions and interaction networks, can be computed.

@@

Our review begins by considering the dynamics of an asexually reproducing population in a homogeneous environment. For simplicity, we first assume individuals are haploid and carry one of $K$ alleles each with a different fitness expressed as growth rate before introducing a model involving a quantitative trait. Under these assumptions, the derivation of the evolution of allele frequencies due to natural selection can be derived from expressions of exponential growth. This, and a few related approaches, have been provided by Crow and Kimura [-@9781932846126, \S 5.3]. Specifically, denoting $\nu_i$ the abundance of individuals with allele $i$ and $m_i$ the growth rate of allele $i$ [called the Malthusian parameter in @9781932846126], we have
\begin{equation}\label{first}
\frac{d\nu_i}{dt}=m_i\nu_i.
\end{equation}
Starting from this model, we get the total abundance of the population as $N=\sum_{i=1}^K\nu_i$, the frequency of allele $i$ as $p_i=\nu_i/N$ and the mean fitness of the population as $\bar m=\sum_{i=1}^Kp_im_i$. Hence, we can employ some elementary calculus to derive the dynamics of abundance $dN/dt$ and the dynamics of allele frequencies $dp_1/dt,\dots,dp_K/dt$ as
\begin{equation}
\frac{dN}{dt}=\sum_{i=1}^K\nu_im_i=N\sum_{i=1}^Kp_im_i=\bar mN,
\end{equation}
\begin{equation}\label{popgen}
\frac{dp_i}{dt}=\frac{d}{dt}\frac{\nu_i}{N}=\frac{1}{N^2}\left(N\frac{d\nu_i}{dt}-\frac{dN}{dt}\nu_i\right)
=\frac{1}{N}\left(m_i\nu_i-\bar m Np_i\right)=(m_i-\bar m)p_i.
\end{equation}
Two important observations of these equations include (i) mean fitness $\bar m$ determines the abundance dynamics of the entire population and (ii) allele $i$ will increase (decrease) in frequency if $m_i>\bar m$ ($<\bar m$). Equation (\ref{popgen}) is known in the field of evolutionary game theory as the replicator equation [@Hofbauer1998; @9780674023383; @Lion2018; @Taylor1978; @Schuster1983]. Instead of being explicitly focused on alleles, the replicator equation describes the fluctuations of relative abundances of various _types_ in a population in terms of the vital rates of each type. Using a matrix of transition rates between differing types, it is straight-forward to extend the replicator equation to include mutation, which is known as the replicator-mutator equation [@9780674023383].

@@

To stay within the realm of biological plausibility we require a set technical assumptions. First, we assume the initial abundance density is continuous, non-negative, integrable and has finite trait mean and variance. That is, we assume $\nu(x,0)$ is continuous in $x$, satisfies $\nu(x,0)\geq0$ for all $x\in\mathbb{R}$ and
\begin{equation}
N(0)=\int_{-\infty}^{+\infty}\nu(x,0)dx<+\infty,
\end{equation}
\begin{equation}
-\infty<\bar x(0)=\int_{-\infty}^{+\infty}xp(x,0)dx<+\infty,
\end{equation}
\begin{equation}
\sigma^2(0)=\int_{-\infty}^{+\infty}(x-\bar x(0))^2p(x,0)dx<+\infty,
\end{equation}
where $\bar x(t)$ and $\sigma^2(t)$ are respectively the mean trait and phenotypic variance at time $t\geq0$. Second, we assume selection is determined by the growth rate $m(h,x)$ that is differentiable with respect to both arguments and is bounded above by some $r\in\mathbb{R}$ for all $x\in\mathbb{R}$ and $h\geq0$. We also assume $(1+h)C\leq m(h,x)$ for some constant $C>0$ so that $m(h,x)$ is linear or more than linear in $h$. We denote by $K$ an operator that takes functions such as $\nu(x,t)$ as its argument and returns other functions that may still depend on the spatial variable $x$ and the temporal variable $t$. For our application we will be focused on operators $K$ that can be written as $(K\nu)(x,t)=\int_{-\infty}^{+\infty}\kappa(x-y)\nu(y,t)dy$ for some non-negative and bounded function $\kappa$. We set $h=(K\nu)(x,t)$ to account for nonlocal effects of the abundance density $\nu(x,t)$ on the fitness of individuals with trait value $x$ [@Volpert2014]. As an example of a nonlocal effect, consider a population experiencing intraspecific competition for a common resource such that the competitive interactions are not mediated by any set of traits. In this case all individuals in the population effect the fitness of one another even though they may exhibit a diverse range of trait values. The term nonlocal comes from interpreting the trait value as spatial location. From here on we abbreviate $m(h,x)=m\big((K\nu)(x,t),x\big)$ to just $m(\nu,x)$. Third, we assume mutation is captured by diffusion with coefficient $\frac{\mu}{2}$. With these technicalities addressed we model the demographic dynamics of a population and the dynamics of a quantitative character simultaneously by the PDE
\begin{equation}\label{eq1}
\frac{\partial}{\partial t}\nu(x,t)=m(\nu,x)\nu(x,t)+\frac{\mu}{2}\frac{\partial^2}{\partial x^2}\nu(x,t)
\end{equation}
with the initial condition $\nu(x,0)$ described above. Equation (\ref{eq1}) is a semilinear parabolic PDE and is referred to as a scalar reaction-diffusion equation [@zheng2004nonlinear; @lawrenceevans2010]. To ensure existence and uniqueness of solutions to (\ref{eq1}), we further assume that for each $M>0$ there exists a constant $L_M>0$ depending on $M$ such that $\int_\mathbb{R}\nu_1(x)dx,\int_\mathbb{R}\nu_2(x)dx\leq M$ implies
\begin{equation}\label{local_lipschitz}
\int_\mathbb{R}\left|m(\nu_1,x)\nu_1(x)-m(\nu_2,x)\nu_2(x)\right|dx\leq L_M\int_\mathbb{R}|\nu_1(x)-\nu_2(x)|dx.
\end{equation}
In SM \S\ref{finite} we combine the above assumptions on growth rate, mutation and initial condition to prove that solutions to (\ref{eq1}) satisfy $N(t),|\bar x(t)|,\sigma^2(t)<+\infty$ for all $t\geq0$.

Interpreting the trait value $x$ as spatial location and ignoring nonlocal effects, which can be satisfied by setting $K\nu=\nu$, this model of spatially distributed population dynamics has been intensely studied [for a review, see @Cantrell2004]. When $\mu=0$, equation (\ref{eq1}) can be seen as an analog of equation (\ref{first}) for a continuum of types. By assuming mutation acts as diffusion, the effect of mutation causes $\nu(x,t)$ to flatten out over time. In fact, if the growth rate is constant across $x$, then this model of mutation will cause $\nu(x,t)$ to converge to a flat line as $t\to\infty$.

Although clearly an idealized representation of biological reality, this model is sufficiently general to capture a large class of dynamics including density dependent growth and frequency dependent selection. As an example, logistic growth combined with quadratic stabilizing selection can be captured using the growth rate 
\begin{equation}\label{special_m}
m(\nu,x)=r-\frac{a}{2}(\theta-x)^2-c\int_{-\infty}^{+\infty}\nu(y,t)dy
\end{equation}
where $r\in\mathbb{R}$ is the intrinsic rate of growth in the absence of abiotic stabilizing selection, $a>0$ is the strength of abiotic stabilizing selection around the phenotypic optimum $\theta\in\mathbb{R}$, $c>0$ captures the sensitivity of fitness to intraspecific competition and $\kappa(x-y)=1$ so that competitive interactions cause the same reduction in fitness regardless of trait value. This exemplary fitness function has a few convenient properties. First, the effect of competition induces a local carrying capacity on the population, leading to a finite equilibrium abundance over bounded subsets of $\mathbb{R}$. Second, abiotic selection prevents the abundance density from diffusing too far from the abiotic optimum, leading to a finite equilibrial abundance across all of $\mathbb{R}$, $\lim_{t\to\infty}N(t)=\tfrac{1}{c}(r-\tfrac{1}{2}\sqrt{a\mu})$, finite equilibrial mean trait $\lim_{t\to\infty}\bar x(t)=\theta$ and finite equilibrial phenotypic variance $\lim_{t\to\infty}\sigma^2(t)=\sqrt{\frac{\mu}{a}}$, so long as $\bar x(0)\in\mathbb{R}$, $\sigma^2(0)\in[0,+\infty)$, $N(0)\in(0,+\infty)$ and $r>\frac{1}{2}\sqrt{a\mu}>0$. We prove these statements in SM \S\ref{equilib}. Note in particular the coincidence of equilibrial phenotypic variance predicted by this model and that predicted by the classic Gaussian allelic model of quantitative genetics [@Lande1975; @9780471986539; @Johnson2005]. In \S\ref{coev} and SM \S\ref{diffuse} we find this fitness function is closely related to a fitness function derived from niche theory that we use to model coevolution driven by resource competition.
